Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-03-09 paper/paper.tex

The most studied version of the above problems is finding the maximum/minimum set size and one of the solution with that size satisfying the contraints,
while many of them belong to the hardness category NP-hard~\cite{Hastad1996}, i.e. there is unlikely a polynomial time algorithm for solving them.
In this paper, we focus on the less studied \textit{property} computation of these problems which can be harder but is crutial towards understanding a problem.
The property computation includes finding the maximum/minimum set size, counting sets of a give size, enumerating all such sets or direct sampling them when they are not enumeratable.
For the weighted version, it also includes finding largest/smallest $k$ sets and their sizes.
%are many interesting and hard computational problems concerning various properties of independent sets.

We provide an abstract algebra view for the computation of all these properties and proposed a unified framework,
generic tensor-network, to solve them all.
%We map the computation of these properties into generic tensor network contraction with specially designed tensor element algebra.
Tensor network is also known as the sum product network or einsum in different fields,
where its tensor element types are typically restricted to standard number types such as real numbers and complex numbers.
While generic tensor network is a generalization of which by extending the tensor element type to a generic one being a commutative semiring.
%Traditional algorithms for finding the MIS size include the branching algorithms~\cite{Tarjan1977, Robson1986} and dynamic programming based algorithms~\cite{Courcelle1990, Fomin2013}.
Given a good contraction order, contracting a tensor network in general has time complexity $2^{{\rm tw}(G)}$, where ${\rm tw}(G)$ is the treewidth of graph $G$.
Recent progress in simulating quantum circuits with random tensor networks~\cite{Gray2021, Pan2021, Kalachev2021} enables us to find the optimal contraction order of a large scale tensor network in reasonable time.
Although this time complexity is just on par with the dynamic programming algorithms~\cite{Courcelle1990, Fomin2013} for solving the maximum size and one of the best set, tensor network has richer algebraic structure, and this structure is crutial for computing properties.
%For sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.

@@

\section{Tensor network representation of independent sets} \label{sec:tnmap}
Let $G=(V,E)$ be a weighted graph with each vertex $i\in V$ being associated with a weight $w_i$.
We map a vertex $i\in V$ to a label $s_i \in \{0, 1\}$ of dimension $2$ in a tensor network, where we use $0$ ($1$) to denote a vertex is absent (present) in the set.
For each label $s_i$, we defined a parametrized rank-one vertex tensor $W(x_i, w_i)$ as
\begin{equation}
    W(x_i, w_i) = \left(\begin{matrix}
        1 \\
        x_i^{w_i}
    \end{matrix}\right).
\end{equation}
We use subscripts to index tensor elements, e.g.\ $W(x_i, w_i)_0=1$ is the first element associated with $s_i=0$ and $W(x_i, w_i)_1=x_i^{w_i}$ is the second element associated with $s_i=1$.
Similarly, on each edge $(u, v)$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction gives
\begin{equation}\label{eq:idp}
    P(G, \{x_1, x_{2}, \ldots,x_{|V|}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{i=1}^{|V|} W(x_i, w_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$ (see Example \ref{eg:tensorcontraction} for a concrete example).
The edge tensor element $B_{s_{i}=1, s_{j}=1} = 0$ encodes the independent set constraint, meaning vertex $i$ and $j$ cannot be both in the independent set if they are connected by an edge $(i,j)$.

@@

>>>>

The most studied version of the above problems is finding the maximum/minimum set size and one of the solution with that size satisfying the contraints,
while many of them belong to the hardness category NP-hard~\cite{Hastad1996}, i.e. there is unlikely a polynomial time algorithm for solving them.
In this paper, we focus on the less studied \textit{property} computation of these problems which can be harder but is crutial towards understanding a problem.
Here, the \textit{property} refers to a class of quantities that not only include the maximum/minimum set size,
but also include the number of sets at a give size, enumeration of all sets at a give size or direct sampling of such sets when they are not enumeratable.
For the weighted version, it also includes finding largest/smallest $k$ sets and their sizes.
%are many interesting and hard computational problems concerning various properties of independent sets.

We propose to view the computation of all these properties from an abstract algebra perspective and use a unified framework,
generic tensor-network, to solve these seeming unrelated properties all.
%We map the computation of these properties into generic tensor network contraction with specially designed tensor element algebra.
Tensor network is also known as the sum product network or einsum in different fields,
where its tensor element types are typically restricted to standard number types such as real numbers and complex numbers.
We extend this concept to generic tensor network by generalizing the tensor element type to any type being a commutative semiring.
%Traditional algorithms for finding the MIS size include the branching algorithms~\cite{Tarjan1977, Robson1986} and dynamic programming based algorithms~\cite{Courcelle1990, Fomin2013}.
Given a good contraction order, contracting a tensor network in general has time complexity $2^{{\rm tw}(G)}$, where ${\rm tw}(G)$ is the treewidth of graph $G$.
Recent progress in simulating quantum circuits with random tensor networks~\cite{Gray2021, Pan2021, Kalachev2021} enables us to find such a good contraction order for a tensor network with thousands of tensors in a reasonable time.
Although this time complexity is just on par with the dynamic programming algorithms~\cite{Courcelle1990, Fomin2013} for solving the maximum size and one of the best set, tensor network has richer algebraic structure, and this structure is crutial for computing properties.
%For sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.

@@

\section{Tensor network representation of independent sets} \label{sec:tnmap}
This section is about reducing the independent set problem to a tensor network contraction problem,
for people with physics background, it is suggested to use an alternative approach described in \App{sec:energymodel} to understand this construction and the independence polynomial from the partition function perspective.
Let $G=(V,E)$ be a weighted graph with each vertex $i\in V$ being associated with a weight $w_i$.
We map a vertex $i\in V$ to a label $s_i \in \{0, 1\}$ of dimension $2$ in a tensor network, where we use $0$ ($1$) to denote a vertex is absent (present) in the set.
For each label $s_i$, we defined a parametrized rank-one vertex tensor $W(x_i, w_i)$ as
\begin{equation}
    W(x_i, w_i) = \left(\begin{matrix}
        1 \\
        x_i^{w_i}
    \end{matrix}\right).\label{eq:vertextensor}
\end{equation}
We use subscripts to index tensor elements, e.g.\ $W(x_i, w_i)_0=1$ is the first element associated with $s_i=0$ and $W(x_i, w_i)_1=x_i^{w_i}$ is the second element associated with $s_i=1$.
Similarly, on each edge $(u, v)$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction gives
\begin{equation}\label{eq:idp}
    P(G, \{x_1, x_{2}, \ldots,x_{|V|}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{i=1}^{|V|} W(x_i, w_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$ (see Example \ref{eg:tensorcontraction} for a concrete example).
The edge tensor element $B_{s_{i}=1, s_{j}=1} = 0$ encodes the independent set constraint, meaning vertex $i$ and $j$ cannot be both in the independent set if they are connected by an edge $(i,j)$.

@@

\section{An alternative way to construct the tensor network}\label{sec:energymodel}

Let us characterize the independent set problem on graph $G=(V, E)$ as an energy model
\begin{equation}
    \mathcal{E}(G, s) = -\sum_{i\in V(G)} w_i s_i + \infty \sum_{\langle i,j\rangle \in E(G)}s_i s_j
\end{equation}
where $s_i$ is a spin on vertex $i \in V$ and $w_i$ is an onsite energy term.
The first term is the energy that corresponds to the inverse of the independent set size and the second term describes the independence contraint.
In physical systems, this contraint term usually corresponds to the Rydberg blockade~\cite{Pichler2018, Ebadi2022} in cold atom arrays or the repulsive force in hard core lattice model~\cite{Dyre2016, Fernandes2007}.
Its partition function is defined as
\begin{equation}
    \begin{split}
    Z(G, \beta) = \sum_{s}e^{-\beta \mathcal{E}(G, s)} = \sum_{s\in \mathcal{I}(G)} e^{\beta \sum w_i s_i}\\
         = \sum_{k=0}^{\alpha(G)}D(k) e^{\beta k}  \qquad \quad (w_i = 1)
    \end{split}
\end{equation}
where $\mathcal{I}(G)$ is the set of independent sets of graph $G$.
The second line corresponds to the unweighted independent set problem,
$\alpha(G)$ is the absolute value of the minimum energy (maximum independent set size) and $D(k)$ is the number of states with energy $-k$ (independent sets of size $k$).
It can be evaluated as a tensor network contraction,
where the tensor network is constructed by placing a vertex tensor on each vertex $i$
\begin{equation}
    W(\beta, w_i) = \left(\begin{matrix}
        1 \\
        e^{\beta w_i}
    \end{matrix}\right),
\end{equation}
and a bond tensor on each edge
\begin{equation}
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right),
\end{equation}
where the $0$ in the bond tensor comes from $e^{-\beta\infty}$ from the second term of the energy model.
By letting $x = e^{\beta w_i}$, we get the tensor network for computing the independence polynomial as described by \Eq{eq:vertextensor} and \Eq{eq:edgetensor}.
